Formalization of Abstract Rewriting

### 2025-02-22 Relations/Paper/Definitions.tex
update after call

\greyout{We denote by $\tilde R$ the relational converse of $R$ ($(x,y) \in \tilde R \iff (y,x) \in R$),
and by $R;R'$ the relational composition of $R$ and $R'$.}

\subsection{Closure operators (\texttt{ClosureOperators.agda})}
\begin{notation}
  The following notions are standard.  For precise definitions,
  see \texttt{ClosureOperators.agda}.
  \begin{enumerate}
    \item $R^r$ denotes the \emph{reflexive closure} of $R$.
    \item $R^s$ denotes the \emph{symmetric closure} of $R$.
    \item $R^+$ denotes the \emph{transitive closure} of $R$.
    \item $R^* = (R^r)^+$ denotes the \emph{reflexive-transitive closure} of $R$.
    \item $R^= = (R^s)^*$ denotes the \emph{equivalence relation generated by} $R$.
  \end{enumerate}
\end{notation}

@@

We next define rewriting-theoretic concepts like confluence and normalization relative to a fixed element of $A$; global versions of these properties are obtained by universal quantification.
\subsection{Normalization (\texttt{ARS-Properties.agda})}
\begin{definition} The following definitions relate to normalization. Let $A$ and $R$ be as above, and let $a \in A$.
  \begin{description}
    \item[$a \in \NF_R$] \emph{$a$ is a normal form} if $(a,b) \notin R$ for any $b \in A$ 
    \item[$a \in \WN_R$] \emph{$a$ is weakly normalizing} if $a \mstep b$ for some $b \in \NF_R$.
    \item[$a \in \SN_R$]  \emph{$a$ is strongly normalizing} if every reduction sequence starting from $a$ is well-founded\footnote{See Section \ref{sec:Well-foundedness} for definitions of being well-founded. Note that we are 
    making use of the alternative definition of $\SN$ provided in \terese.}. \sacomment{Check this defn}
  \end{description}
\end{definition}

\begin{definition} The folowing definitions relate to being minimal. Let $A$ and $R$ be as above, and let $a \in A$.
    \begin{description}
        \item[$a \in \MF_R$] \emph{$a$ is a minimal form} if $a \mstep b \implies b \mstep a$ for any $b \in A$.
        \item[$a \in \WM_R$] \emph{$a$ is weakly minimalizing} if $a \mstep b$ for some $b \in \MF_R$.
        \item[$a \in \SM_R$]  \emph{$a$ is strongly minimalizing} if either $a \in MF_R$, or every element one $R$ step from $a$ is also strongly minimalizing.
    \end{description}
\end{definition}
We found that there was a strong interrelationship between the normalizing properties and the minimal properties. This relationship is 
captured in Figure \ref{fig:norm-hierarchy}.

\begin{figure}[h]   
    \centering
        \begin{tikzpicture}[auto,
            arrowstyle/.style={->, line width=1pt, >={Latex[length=3mm, width=2mm]}, shorten >=2pt}]
            % A box style
            \tikzstyle{boxnode} = [draw, rectangle, text centered,minimum width=1.2cm, inner sep=3pt]   
            % Define the nodes
            \node (SN) [boxnode] at (4,0) {$\SN$};
            \node (WN) [boxnode, above right=0.5cm and 0.5cm of SN] {$\WN$};
            \node (SR) [boxnode, below right=0.5cm and 0.5cm of SN] {$\SM$};
            \node (WR) [boxnode, below right=0.5cm and 0.5cm of WN] {$\WM$};

            % Draw the arrows to form a diamond
            \draw [arrowstyle] (SN) -- (WN);
            \draw [arrowstyle] (SN) -- (SR);
            \draw [arrowstyle] (WN) -- (WR);
            \draw [arrowstyle] (SR) -- (WR);
        \end{tikzpicture}
        \caption{A hierarchy of normalizing and minimal properties.}
        \label{fig:norm-hierarchy}
\end{figure}

The implications will be explored in detail in Section \ref{sec:Implications}.

A trivial implication not captured in the above figure is covered in Proposition \ref{prop:nftomf}. 
\begin{proposition}\label{prop:nftomf}
    $a\in NF_R \implies a \in MF_R$ 
\end{proposition}    
\begin{proof}
    See, \verb|NF ⊆ MF : ∀x → NFx → MFx| in \texttt{ARS-Implications.agda}.
\end{proof}
\subsection{Confluence (\texttt{ARS-Properties.agda})}
\begin{definition} The following definitions relate to confluence. Let $A$ and $R$ be as above, and let $a \in A$.
    \begin{description}
        \item[$a \in \WCR_R$] \emph{$a$ is weakly Church-Rosser (or weakly confluent)} if $c \bstep a \rstep b \implies c \mstep d \bmstep b$.
        \item[$a \in \CR_R$] \emph{$a$ is Church-Rosser (or confluent)} if $c \bmstep a \mstep b \implies c \mstep d \bmstep b$.
        \item[$a \in \NPe_R$] \emph{$a$ has the equivalence normal form property} if $a \estep b \implies a\mstep b$ for any $b \in \NF$.
        \item[$a \in \NP_R$] \emph{$a$ has the normal form property} if $c \bmstep a \mstep b \implies c \mstep b$ for any $b \in \NF$\footnote{In \terese $\NF$ is used to denote $\NPe$, whereas we have chosen to use $\NF$ to denote normal form.}. 
        \item[$a \in \MP_R$] \emph{$a$ has the minimal form property} if $c \bmstep a \mstep b \implies c \mstep b$ for any $b \in \MF$.
        \item[$a \in \UN_R$] \emph{$a$ has the unique normal form property} if $a \estep b \implies a \equiv b$  for any $a, b \in NF$.
        \item[$a \in \UNto_R$] \emph{$a$ has the unique normal form property with respect to reduction} if $c \bmstep a \mstep b  \implies b \equiv c$  for any $b, c \in NF$.
    \end{description}
\end{definition}

Again, we found that there was a strong interrelationship between the properties, captured in Figure \ref{fig:conf-hierarchy}, which are explored in Section \ref{sec:Implications}.

\begin{figure}[h] 
    \centering

        \begin{tikzpicture}[auto,
          arrowstyle/.style={->, line width=1pt, >={Latex[length=3mm, width=2mm]}, shorten >=2pt}]
          % A box style
          \tikzstyle{boxnode} = [draw, rectangle, text centered,minimum width=1.2cm, inner sep=3pt]

          % Place the nodes vertically (top to bottom)
          \node (Confluent) [boxnode] at (-4,0) {$\CR$};
          \node (RP)        [boxnode, right=.5cm of Confluent] {$\MP$};
          \node (WN)        [boxnode, right=.5cm of RP] {$\NP$};
          \node (UN)        [boxnode, right=.5cm of WN] {$\UNto$};

          % Draw arrows downwards
          \draw[arrowstyle] (Confluent) -- (RP);
          \draw[arrowstyle] (RP) -- (WN);
          \draw[arrowstyle] (WN) -- (UN);

          \end{tikzpicture}
          \caption{A hierarchy of confluent properties.}
          \label{fig:conf-hierarchy}

\end{figure}

\begin{proposition}
    $\NPe \leftrightarrow \NP$ 
\end{proposition}
\begin{proof}
    See, \verb|NP₌↔NP : R isNP₌ ↔ R isNP| in \texttt{ARS-Implications.agda}.
\end{proof}
\begin{proposition}
    $\UN \implies \UNto $
\end{proposition}
\begin{proof}
    See, \verb|UN→UN→ : R isUN → R isUN→| in \texttt{ARS-Implications.agda}
\end{proof}
The implication $\UNto \implies \UN$ does not hold, as can be seen in CE-\ref{CE:un} in Subsection \ref{subsec:counterexamples}.
\subsection{Sequences and Recurrence (\texttt{ARS-Properties.agda}, \texttt{Seq.agda})}
\begin{definition} The following definitions relate to the minimal definitions and are key to formalizing Theorem 1.2.3 in \terese. Let $A$ and $R$ be as above, and let $a \in A$.
    \begin{description}
        \item[$s : \nat \to A$] \emph{A sequence $s \in A$} is a function from the natural numbers to $A$.  
        \item[$s \in R-increasing$] \emph{A sequence $s \in A$ is $R$-increasing} if every term is $R$-related to its preceding term.
        \item[$a \in \BP$] \emph{$a$ is a bound} if there exists a sequence $s$ such that $s (i) \mstep a$ for any $i \in \nat$. 
        \item[$R\, is\, \RP$] \emph{$R$ has the recurrence property} if every bounded, increasing sequence has an element $a \in MF$.
        \item[$R\, is\, \RPm$] \emph{$R$ has the weak recurrence property} if for every increasing sequence $s$ with bound $b$ there exists an element $i$ of the sequence such that $a \mstep s (i)$.
        \item[$R\, is\, Inc$] \emph{$R$ is increasing} if there exists a ``size function'' $|-| : A \to \nat$ satisfying, for all $a, b \in A$,
        \[ Rab \to |a| < |b| \]       
    \end{description}
\end{definition}
Proposition \ref{prop:SMRP} connects $\SM$ with $\RP$ and $\BP$.
\begin{proposition}\label{prop:SMRP}
    $SM \leftrightarrow RP \land BP$
\end{proposition}
\begin{proof}
    See,\\ \verb|RP∧BP→SMseq : R isRP → R isBP → ∀ {x : A} → SMseq R x| \\ \verb|RisSMseq→RisRP : (∀ {x : A} → SMseq R x) → R isRP|
    \\ \verb|RisSMseq→RisBP : (∀ {x : A} → SMseq R x) → R isBP| in \texttt{ARS-Implications.agda}.
\end{proof}

\begin{proposition}
    If $R$ is increasing, then no infinite $R$-increasing sequence is bounded.
\end{proposition}
\begin{proof}
    Let $(a_i) = a_0 \to a_1 \to a_2 \to \cdots$ be an $R$-increasing sequence
    and suppose that $(a_i)$ is bounded.  Let $b$ be the bound, so that $a_i \to b$ for all $i$.

    If $m = |b|$ then we have
    \[ 0 \le |a_0| < |a_1| < |a_2| < \cdots < |a_m| < |b| = m \]
    % \[ m > |a_m| > |a_{m-1}| > \cdots > |a_0| \ge 0 \]
    which is a contradiction, since there are only $m-1$ integers strictly less than $m$.
\end{proof}
It follows that every relation that is increasing in the sense of \terese
vacuously satisfies the RP condition.

\begin{corollary}
    If $R$ is increasing, then $R$ has the recurrence property.
\end{corollary}
We discuss in Section \ref{sec:Formalization} the motivation for finding an alternative for the increasing property found in \terese.

\sacomment{Anything else to define such as finitely branching?}

>>>>

% \greyout{We denote by $\tilde R$ the relational converse of $R$ ($(x,y) \in \tilde R \iff (y,x) \in R$),
% and by $R;R'$ the relational composition of $R$ and $R'$.}

\subsection{Closure operators (\texttt{ClosureOperators.agda})}
\begin{notation}
  The following notions are standard.  For precise definitions,
  see \texttt{ClosureOperators.agda}.
  \begin{enumerate}
    \item $R^r$ denotes the \emph{reflexive closure} of $R$.
    \item $R^s$ denotes the \emph{symmetric closure} of $R$.
    \item $\tclos{R}$ denotes the \emph{transitive closure} of $R$.
    \item $R^* = (R^r)\tclos{}$ denotes the \emph{reflexive-transitive closure} of $R$.
    \item $R^= = (R^s)^*$ denotes the \emph{equivalence relation generated by} $R$.
  \end{enumerate}
\end{notation}

@@

We next define rewriting-theoretic concepts like confluence and normalization relative to a fixed element of $A$; global versions of these properties are obtained by universal quantification.
\subsection{Normalization (\texttt{ARS-Properties.agda})}
\begin{definition} The following definitions relate to normalization. Let $A$ and $R$ be as above, and let $a \in A$.
  \begin{description}
    \item[$a \in \NF_R$] \emph{$a$ is a normal form} if $(a,b) \notin R$ for any $b \in A$ 
    \item[$a \in \WN_R$] \emph{$a$ is weakly normalizing} if $a \mstep b$ for some $b \in \NF_R$.
    \item[$a \in \SN_R$] \emph{$a$ is strongly normalizing} if $a$ is accessible with respect to the converse relation $\bar{R} = {(y,x) | (x,y) \in R}$. See Section \ref{sec:Well-foundedness}. 
  \end{description}
\end{definition}

\subsection{Confluence (\texttt{ARS-Properties.agda})}
\begin{definition} The following definitions relate to confluence. Let $A$ and $R$ be as above, and let $a \in A$.
    \begin{description}
        \item[$a \in \WCR_R$] \emph{$a$ is weakly Church-Rosser (or weakly confluent)} if $c \bstep a \rstep b \implies c \mstep d \bmstep b$.
        \item[$a \in \CR_R$] \emph{$a$ is Church-Rosser (or confluent)} if $c \bmstep a \mstep b \implies c \mstep d \bmstep b$.
        \item[$a \in \NPe_R$] \emph{$a$ has the equivalence normal form property} if $a \estep b \implies a\mstep b$ for any $b \in \NF$.
        \item[$a \in \NP_R$] \emph{$a$ has the normal form property} if $c \bmstep a \mstep b \implies c \mstep b$ for any $b \in \NF$. 
        \item[$a \in \UN_R$] \emph{$a$ has the unique normal form property} if $a \estep b \implies a \equiv b$  for any $a, b \in \NF$.
        \item[$a \in \UNto_R$] \emph{$a$ has the unique normal form property with respect to reduction} if $c \bmstep a \mstep b  \implies b \equiv c$  for any $b, c \in \NF$.
    \end{description}
\end{definition}
For each of the subsets defined above we denote in bold face the proposition that the subset contains all of $A$.
That is, for $\mathrm{X} \in \{\WCR , \CR , \NP , \UN \}$ we write $\mathbf{X}$ for the statement that $x \in \mathrm{X}_R$ for all $x \in A$.

In \terese $\NF$ is used to denote $\NPe$, whereas we have chosen to use $\NF$ to denote normal form. Moreover, ${\gNP}$ and ${\gNPe}$ are equivalent when

\begin{proposition}
    $\NPe \leftrightarrow \NP$ 
\end{proposition}
\begin{proof}
    See, \verb|NP₌↔NP : R isNP₌ ↔ R isNP| in \texttt{ARS-Implications.agda}.
\end{proof}
\begin{proposition}
    $\UN \implies \UNto $
\end{proposition}
\begin{proof}
    See, \verb|UN→UN→ : R isUN → R isUN→| in \texttt{ARS-Implications.agda}
\end{proof}
The implication $\UNto \implies \UN$ does not hold, as can be seen in CE-\ref{CE:un} in Subsection \ref{subsec:counterexamples}.
\subsection{Sequences and Recurrence (\texttt{ARS-Properties.agda}, \texttt{Seq.agda})}
The following definitions are key to formalizing Theorem 1.2.3 in \terese.
$R \models \mathbf{X}$ is $\forall a. a \in \mathrm{X}_R$. When $R$ is clear from the context, we will write $\gNP$ for $R \models \gNP$, etc. \sacomment{Mention this earlier}

\begin{definition}
    \begin{description}
        \item[$R \models Inc$] \emph{$R$ is increasing} if there exists a ``size function'' $|-| : A \to \nat$ satisfying, for all $a, b \in A$,
        $Rab \to |a| < |b|$  
    \end{description}
\end{definition}

\sacomment{motivate here going from above defn to the following.}

\begin{definition} Let $A$ and $R$ be as above, and let $a \in A$.
    \begin{description}
        \item[$s : \nat \to A$] \emph{A sequence $s \in A$} is a function from the natural numbers to $A$.  
        \item[$s \in R-increasing$] \emph{A sequence $s \in A$ is $R$-increasing} if every term is $R$-related to its preceding term.
        \item[$a$ is an $s$-bound] if $s (i) \mstep a$ for all $i \in \nat$. 
        \item[$a \in \BP_R$] if all $R$-increasing sequence $s$ with $s (0) = a$ there exists an $s$-bound. 
        \item[$R \models \RPm$] \emph{$R$ has the weak recurrence property} if for every increasing sequence $s$ with bound $b$ there exists an $i \in \nat$ such that $b \mstep s (i)$.
        \item[$R \models \RP$] \emph{$R$ has the recurrence property} if every bounded, increasing sequence $s$ with an $s$-bound, there exists an $i \in \nat$ such that 
        $s (i) \in \MF$.
    \end{description}
\end{definition}

The following is immediate.
\begin{proposition}\hfill 
    \begin{enumerate}
        \item $\Inc \implies \RPm$
        \item $\RPm \implies \RP$
    \end{enumerate}
\end{proposition}

It follows that every relation that is increasing in the sense of \terese
vacuously satisfies the RP condition.

\sacomment{tidy up below on rp iff rp- .}
We investigated $\RP$ and thought that with $\RPm$ we had a candidate for a weaker definition that 
did not rely on the notion of $\MF$. We were able to proove that this weaker definition is 
in fact an equivalent definition in the following proof in Agda:

\verb|RP-↔RP : R is RP- ↔ R is RP| \footnotemark[1]

\footnotetext[1]{\texttt{ARS-implications.agda}}
\subsection{Finitely branching}
\sacomment{This property is what makes a lot of the classical properties valid which otherwise wouldn't be. Valid in most interesting rewriting systems. Most 
programming languages will satisfy finite branching.}
\begin{definition}
    \begin{description}
        \item[TODO] TODO 
    \end{description}
\end{definition}
\sacomment{Anything else to define such as finitely branching?}

### 2025-02-22 Relations/Paper/Formalization.tex
update after call

\section{Formalization of Terese}
\label{sec:Formalization}
This sections acts as a guide of our formalization of \terese and explains the deviations we took.

@@

Our motivation for using $\RP$ rather than $\Inc$ is that we found $\Inc$ to be a particularly difficult property 
to use in a constructive setting.

\subsection{Propositions}
There are three sets of propositions given in \terese: \texttt{1.1.9}, \texttt{1.1.10}, \texttt{1.1.11}. 
The formalization of the proofs are in \texttt{ARS-Propositions.agda}. These formalizations were straightforward 
and as such are not the focus of this paper.

@@

We investigated $\RP$ and thought that with $\RPm$ we had a candidate for a weaker definition that 
did not rely on the notion of $\MF$. We were able to proove that this weaker definition is 
in fact an equivalent definition in the following proof in Agda:

\verb|RP-↔RP : R is RP- ↔ R is RP| \footnotemark[1]
\footnotetext[1]{\texttt{ARS-implications.agda}}
Three proofs are given in \terese of Newman's Lemma (Theorem 1.2.1). The first and third of 
these proofs we found to rely on classical principles. The second proof was amenablbe to a 
constructive formalization which we carried out in the function

>>>>

\section{Formalization of \terese}
\label{sec:Formalization}
This sections acts as a guide of our formalization of \terese and explains the deviations we took.

@@

Our motivation for using $\RP$ rather than $\Inc$ is that we found $\Inc$ to be a particularly difficult property 
to use in a constructive setting. \sacomment{Say more on this motivation and decide where it will be placed.}

\subsection{Propositions}
There are three sets of propositions given in \terese: \texttt{1.1.9}, \texttt{1.1.10}, \texttt{1.1.11}. 
The formalization of the proofs are in \texttt{ARS-Propositions.agda}. These formalizations were straightforward 
and as such are not the focus of this paper. \sacomment{Do we include more information on these propositions?}

@@

Three proofs are given in \terese of Newman's Lemma (Theorem 1.2.1). The first and third of 
these proofs we found to rely on classical principles. The second proof was amenablbe to a 
constructive formalization which we carried out in the function

### 2025-02-22 Relations/Paper/Implications.tex
update after call

>>>>

\begin{definition} The folowing definitions relate to being minimal. Let $A$ and $R$ be as above, and let $a \in A$.
    \begin{description}
        \item[$a \in \MF_R$] \emph{$a$ is a minimal form} if $a \mstep b \implies b \mstep a$ for any $b \in A$.
        \item[$a \in \WM_R$] \emph{$a$ is weakly minimalizing} if $a \mstep b$ for some $b \in \MF_R$.
        \item[$a \in \SM_R$]  \emph{$a$ is strongly minimalizing} if either $a \in MF_R$, or every element one $R$ step from $a$ is also strongly minimalizing.
        \item[$a \in \MP_R$] \emph{$a$ has the minimal form property} if $c \bmstep a \mstep b \implies c \mstep b$ for any $b \in \MF$.

    \end{description}
\end{definition}

Again, we found that there was a strong interrelationship between the properties, captured in Figure \ref{fig:conf-hierarchy}, which are explored in Section \ref{sec:Implications}.

\begin{figure}[h] 
    \centering

          \end{tikzpicture}
          \caption{A hierarchy of confluent properties.}
          \label{fig:conf-hierarchy}

\end{figure}
We found that there was a strong interrelationship between the normalizing properties and the minimal properties. This relationship is 
captured in Figure \ref{fig:norm-hierarchy}.

\begin{figure}[h]   
    \centering
        \begin{tikzpicture}[auto,
            arrowstyle/.style={->, line width=1pt, >={Latex[length=3mm, width=2mm]}, shorten >=2pt}]
            % A box style
            \tikzstyle{boxnode} = [draw, rectangle, text centered,minimum width=1.2cm, inner sep=3pt]   
            % Define the nodes
            \node (SN) [boxnode] at (4,0) {$\SN$};
            \node (WN) [boxnode, above right=0.5cm and 0.5cm of SN] {$\WN$};
            \node (SR) [boxnode, below right=0.5cm and 0.5cm of SN] {$\SM$};
            \node (WR) [boxnode, below right=0.5cm and 0.5cm of WN] {$\WM$};

            % Draw the arrows to form a diamond
            \draw [arrowstyle] (SN) -- (WN);
            \draw [arrowstyle] (SN) -- (SR);
            \draw [arrowstyle] (WN) -- (WR);
            \draw [arrowstyle] (SR) -- (WR);
        \end{tikzpicture}
        \caption{A hierarchy of normalizing and minimal properties.}
        \label{fig:norm-hierarchy}
\end{figure}

The implications will be explored in detail in Section \ref{sec:Implications}.

Proposition \ref{prop:SMRP} connects $\SM$ with $\RP$ and $\BP$. \sacomment{Need to relate SM and SMseq. Probably define SMseq and metion 
we can only prove this for SMseq.}
\begin{proposition}\label{prop:SMRP}
    $SM \leftrightarrow RP \land BP$
\end{proposition}
\begin{proof}
    See,\\ \verb|RP∧BP→SMseq : R isRP → R isBP → ∀ {x : A} → SMseq R x| \\ \verb|RisSMseq→RisRP : (∀ {x : A} → SMseq R x) → R isRP|
    \\ \verb|RisSMseq→RisBP : (∀ {x : A} → SMseq R x) → R isBP| in \texttt{ARS-Implications.agda}.
\end{proof}

A trivial implication not captured in the above figure is covered in Proposition \ref{prop:nftomf}. 
\begin{proposition}\label{prop:nftomf}
    $a\in NF_R \implies a \in MF_R$ 
\end{proposition}    
\begin{proof}
    See, \verb|NF ⊆ MF : ∀x → NFx → MFx| in \texttt{ARS-Implications.agda}.
\end{proof}
